Inferring causal connectivity from pairwise recordings and optogenetics

### 2018-11-01 manuscript/main.tex
update manuscript to latest

\documentclass[11pt]{article}
\usepackage{amsmath,array}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{breakcites}
\usepackage{graphicx,subcaption}
\usepackage{todonotes} %[disable]
\usepackage{cleveref}
\usepackage{tikz}
\usepackage{placeins} %FloatBarrier
\usepackage{booktabs}
\usepackage{natbib}
\usepackage{authblk}
\usepackage{url}

\usetikzlibrary{arrows,shapes,calc}

@@

 \maketitle
% suggestions for referee?
\begin{abstract} % max words 150
\noindent To study how the brain works, it is essential to identify causal interactions between neurons, usually requiring perturbations. However, when using optogenetics we typically perturb multiple cells, producing a confound - any of the stimulated neurons can have affected the target neuron. Here we show how this produces large biases, and how they can be reduced using the instrumental variable (IV) framework from econometrics. The interaction between stimulation and the absolute refractory period produces a weak, approximately random signal which can be exploited to estimate causality. When simulating integrate-and-fire neurons, we find that IV estimates connectivity better than na\"ive techniques ($R^2=0.845$ vs $R^2=0.048$). The difference is important as the estimates differ when applied to experimental data from stimulated neurons with recorded spiking activity. Our IV approach is easy to implement in the context of current experimental paradigms.

@@

\section{Introduction}
% The brain is infinitely confounded, thus correlations are not sufficient to draw conclusions about mechanisms and this is a hard problem in brain data.
A goal of neuroscience is to understand the mechanisms or causal chains that give rise to activity in the brain, to perception, cognition, and action. 
Complex systems such as the brain are hard to understand because of the numerous ways the contributing elements may interact internally \citep{Jonas2017}. 
Because of this, it is not sufficient to know the correlations between variables or even be able to predict them.
While observing correlations within the system is relatively easy, transitioning from observed correlations to a causal or mechanistic understanding is hard.
After all, there can be many ways that the same activities emerge from distinct causal chains \citep{drton2011global, peters2017elements}. 
Reaching a mechanistic level of understanding in the human brain is nearly impossible as it contains approximately 86 billion neurons \citep{azevedo2009equal}, each of which influences many other neurons.
Even if we could record all neurons at the same time, estimating causality and producing a mechanistic understanding would be extremely challenging.

% We use observational studies, where identifiability is questionable and causality is thus typically wrong. 
Typically, we only record from a small subset of all neurons. 
The data we obtain from such recordings, e.g. from electrophysiology or calcium imaging, is observational, which means that it does not result from randomized perturbations. 
In such cases, we can never know to which level the observed activity was caused by other observed activity, or by unobserved activity. 
The unobserved neurons are called confounders.
If mechanisms are estimated from observational data in the presence of confounders, the consequence may be large errors and incorrect conclusions \citep{angrist2008mostly}. 
Unobserved neural activity confounds estimates of causal interactions and makes it difficult to estimate underlying mechanisms.

% Set up confounding
Confounding is the big threat to causal validity \citep{UpcomingMehlerPaper} irrespective of the use of simple regression techniques or advanced functional connectivity techniques \citep{stevenson2008inferring, honey2009predicting, aitchison2017or, pfau2013robust}.  
To estimate connectivity it is first and foremost important that it stems from cause and effect, therefore we use the term causal connectivity. 
Consider an example where we want to estimate causal connectivity between two observed, but unconnected neurons, $ A $ and $ C $ (\cref{fig:intro}(a)). 
A third unobserved neuron $ B $, is driven together with $ A $ by a common input and they are thus strongly correlated.
Furthermore, $ B $ and $ C $ are connected, thus the input drives $ C $ through $ B $.
Consequently, $ A $ and $ C $ are also correlated and the regression $ C = \beta A $ will misleadingly conclude a direct interaction when causally interpreted. 
In this case, we say the regressor $ A $ is endogenous and the regression coefficient $ \beta $ estimates the magnitude of association rather than the magnitude of causation. 
Na\"ive regressions in partially observed systems will generally not reveal causality.

% Why do we not use GLM
A much used method for estimating the output of single neurons is to perform multiple regression analyses \citep{pillow2008spatio}, modeling each neuron with a generalized linear model (GLM). 
Multiple regression may be a solution to confounding problems as they support ``explaining away" background activity \citep{stevenson2008inferring}.
However, this is only a good strategy if most neurons are included in the recordings.
Furthermore, only under certain assumptions about nonlinearity or noise sources does a fully observed system become identifiable \citep{daniusis2012inferring,shimizu2006linear}.
This is rarely the case in experimental settings, especially in the mammalian brain.
Thus, brain data, will almost never satisfy the criteria needed for identifiability \citep{pearl2009causality}. 
We thus aim to develop methods that can identify causal connectivity between neuron pairs.

% Stimulation is an intervention which is needed to estimate causal interactions
To estimate causal relationships between neurons, stimulating the pre-synaptic neuron is the gold standard. 
In fact, a common definition of causality is in terms of the effect of changing one variable in the system, independently of changing other variables -- an intervention \citep{pearl2009causality}. 
If we stimulate single neurons, the ability to estimate causal relationships by regression is within reach. 
However, this is experimentally challenging and yield low cell count because it requires intracellular, juxtacellular or two-photon stimulation \citep{pinault1996novel, lerman2017two, nikolenko2007two, emiliani2015all}. 
Because gold-standard perturbations are challenging, it is necessary and highly desirable if causality could be obtained from optogenetic stimulation in combination with neural recordings of large populations of neurons \citep{boyden2005millisecond, zemelman2002selective}.

% Optogenetic stimulation is usually totally non-local, producing confounders
Interpreting the results from optogenetic stimulation in terms of causal interactions is difficult. 
In most experimental settings, optogenetic stimulation will affect many neurons simultaneously. 
Hence, the stimulus will produce a distributed pattern of activity.
This distributed pattern of stimulation produces activity which then percolates through the network of neurons. 
Thus any post-synaptic activity induced by stimulation could in principle come from any of the stimulated neurons introducing problematic confounders.

% Causal inference is a thing, if not done well you say stupid things
For insights into how we may resolve the confounding problem induced by optogenetic stimulation, we may look to other fields that have addressed the problem of endogenous regressors. 
The inference of causality from observational data is addressed in the fields of statistics \citep{pearl2009causality}, machine learning \citep{peters2017elements} and econometrics \citep{angrist2008mostly}.

% Example
A commonly used approach towards causal inference in economics are instrumental variables. 
Let's say that we want to estimate the return $ \beta $ from education $ x $ to yearly wages $ y $ with the regression $ y = \beta x + u $. 
Here $ u $ are the factors other than education that contribute to yearly wages. 
One of the factors in $ u $ is a person's cognitive ability which may also affect education and thus the regressor $ x $  is correlated with the error term $ u $. 
This will imply that the regression estimate $ \beta $ will not estimate the magnitude of causation from education on wages, but rather, its association. 
In this case one may use the proximity to a college or university as an instrumental variable (IV) \citep{card1993using}. 
Proximity is a good instrument since it can only affect wages through schooling. 
We expect living in proximity of colleges and universities to give higher probability to attend without affecting other contributing factors to wages such as cognitive ability. 
Then, in order to attribute the causal effect of education on wages one may calculate the ratio of covariances $ \beta = cov(\mathrm{proximity},\mathrm{wages})/cov(\mathrm{proximity},\mathrm{education})$. 
This ratio corrects for the confounding factor.

% Worry is this really correct? Yes, its provable. IVs are (under the right assumptions) actually correct.
The IV technique has been used extensively in econometrics and is provable causal given three assumptions \citep{angrist2008mostly}. 
First, the instrument must be decorrelated with the error term. 
Second, the instrument must be correlated with the regressor. 
Third, there must be no direct influence of the instrument on the outcome variable, but only an influence through the regressor variable. 
The validity of these assumptions is central when using the IV approach.

% Refractory period as an IV
For an instrument to be good, it needs to be unaffected by other variables. 
In the brain, almost everything is affected by the network state. 
However, certain variables can be more or less affected. 
For example, the overall activity of the network is due to slow and strongly nonrandom dynamics. 
In contrast, the temporal pattern of when a neuron is in a refractory state may be random. 
First, if neurons are spiking according to conditional Poisson distributions, their exact timing conditioned on the network state, will be random. 
While refractoriness may not be perfectly random, the exact times of spiking are notoriously difficult to predict \citep{stevenson2008inferring} suggesting that refractoriness is sufficiently random.

@@

\section{Results}
\subsection{Optogenetics is not local}
Optogenetic stimulation is generally seen as a perturbation that by and large affects neurons in proximity (< 1mm) of the light source with the effect decreasing with distance. 
However, this may be an incorrect way of conceptualizing the spatial effect of stimulation as it depends on multiple factors. 
Light intensity and opsin density are important as more light and ion channels will cause a stronger effect on each cell. 
Moreover, the number of potentially stimulated neurons is critical as more neurons will have a larger impact on the overall population activity. 
Finally, physiological properties of the cells are important as light may have a stronger effect on spiking activity when the membrane potential of the cell is sufficiently close to the firing threshold. 
The induced effect of optogenetic stimulation as a function of distance should be given by a relation between the four parameters: light intensity, spatial distribution of neurons, distributions of membrane potential across neurons, and the distribution of induced photo current.

To estimate the light intensity, we calculated the spatial extent of laser light delivered by fiber-optics under plausible experimental conditions according to \citet{Aravanis2007}; see \cref{sec:method:opto}. 
This modeling of light intensity yield an approximately $ 1/r^2 $ reduction with distance $ r $ from the stimulation site \cref{fig:concept} (cyan line). 
This is explained by the surface of a 3D shell growing with $4\pi r^2$ and photons will be roughly isotropic (but see \citep{thunemann2018does}) beyond the scattering length \cref{fig:concept} (inset). 
The same number of photons has to cross each of the spheres around the stimulation location unless they are absorbed. 
As a result, the density of photons decreases rapidly with distance.

The number of illuminated neurons at a given distance will increase with distance to the stimulation site given that neurons are uniformly distributed in brain tissue \cref{fig:concept} (black line). 
In fact, it will increase by approximately $ r^2 $ with distance. 
This derives from the same surface scaling as for the 3D shell. 
Thus the number of neurons that can be activated increases rapidly with distance.

To estimate the effect of stimulation, the functionality underlying spiking activity needs to be considered. 
This can largely be characterized by the  distribution of membrane potentials across neurons. 
Surprisingly, this distribution has been observed to be symmetrically distributed and relatively flat \citep{pare1998impact,destexhe1999impact,rudolph2006use}. 
The expected response from a pulse of light that induces a charge $Q$ should be proportional to the number of affected neurons whose membrane potential sit within a $Q/C$ range of the threshold ($C$ is the capacitance). 
Given that the distribution of membrane potentials is relatively flat (the density close to the threshold is generally within an order of magnitude of the density of its mode) suggests that the spiking response to a perturbation for any neuron is roughly proportional to the induced photo current. 
The peak amplitude of the photo current relates approximately logarithmically to the light intensity \citep{wang2007high}; see \cref{sec:method:opto}. 
Assuming that opsins are evenly distributed across neurons, the induced photo current will not be proportional with light intensity - it will fall slower. 
Based on this, we calculate the overall stimulation effect to be the product of the number of neurons in a spherical slice and the peak amplitude photo current. 
This product actually increases with distance (up to the distance where absorption becomes important) \cref{fig:concept} (blue line). 
Thus, optogenetic stimulation utilizing single photon activation does not produce a localized effect.

\begin{figure}\includegraphics[scale=1]{opto-powerlaw}
\caption{{\bf Spatial extent of optogenetic stimulus}. 
Due to scattering and geometric loss the light intensity (I, cyan line) with an intensity of $10 mW/mm^2$ exiting the optogenetic fiber follows approximately an inverse square law $ r^{-2} $ where $ r $ is the distance from the fiber. 
If neurons are uniformly distributed, the number of affected neurons in a spherical slice increases by $ r^{2} $ (N, black line). 
The total photo current (sum over neurons of peak amplitude photo current in a spherical slice) thus increases with distance due to the nonlinear relation between light intensity and photo current (P, blue line) depicted as percentage of maximum.}
\label{fig:concept}
\end{figure}

@@

After running the simulation, the peri-stimulus time histogram of the stimulated neurons (\cref{fig:intro}(b)) shows the result of both the stimulation itself (suppressed for visibility) and the neuron's refractory period \cref{fig:intro}(c) (AA, BB).
Since the stimulation affects A and B simultaneously, it induces a strong correlation between A and B \cref{fig:intro}(c) (AB). 
This further generates a strong correlation between A and C, confounding the system by rendering the cross-correlation histograms (CCHs) between BC and AC both statistically significant ($ p_{\mathrm{fast}} < 0.001, p_{\mathrm{diff}} < 0.001 $; see \cref{sec:method:cch}). 
Even though the correlation peak between B and C is larger than between A and C, due to correlated spikes outside the periods with stimulation, one may imagine a situation where only A and C is measured, giving rise to a false prediction that they are connected. 
If stimulation affects multiple neurons simultaneously, there is a real confounding problem.

\begin{figure} 
\makeatletter
\renewcommand\p@subfigure{}
\makeatother
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{simple}
\caption{} \label{fig:intro:1}
\end{subfigure}\hfill
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{psth_triple}
\caption{} \label{fig:intro:2}
\end{subfigure}\medskip\\
\begin{subfigure}{\textwidth}\centering \includegraphics[scale=1]{xcorr}
\caption{} \label{fig:intro:3}
\end{subfigure}
\caption{{\bf Optogenetic stimulation induces spurious correlations}. 
A sketch of the simple network containing three neurons shows stimulation configuration with blue laser light and the connections with arrows (a). 
The neurons A and B are stimulated in 1000 trials and the corresponding peristimulus time-histogram are shown in (b) upper panel with a raster plot in the lower panel. 
Cross-correlation histograms (CCHs) are shown in (c) where horizontal and vertical axes represent time lag in ms and counts of coincident spikes in bins of 1 ms.}
\label{fig:intro}
\end{figure}

\FloatBarrier
\subsection{Instrumental variables to resolve confounding}
In order to estimate the actual influence of stimulation of a neuron on post-synaptic neurons, we need to distinguish the influence of one stimulated neuron from the influence of another stimulated neuron. 
We would thus need something that affects the stimulation effect separately between neurons. 
Arguably, refractoriness is such a variable. 
If a neuron is in its absolute refractory period, then no amount of stimulation will make it spike. 
This gives us an interesting way of inferring causality, by comparing the network state between a time when a neuron is able to spike and when the neuron is unable to spike.

Instrumental variables require the existence of a random (or sufficiently random) variable that affects the variable of interest. 
This independent influence then allows quantifying the influence of the variable of interest on the rest of the network. 
In our case, the refractory states of a neuron is in good approximation independent on small time scales (see Discussion for caveats). 
It affects the influence of stimulation on the putative pre-synaptic neuron (\cref{fig:cch_iv_triple}(a)). 
The trials where a stimuli is unable to elicit a spike due to the refractory state can then be used to identify causal effect on the putative post-synaptic neuron.

We can now investigate if the use of an instrumental variable gives a better estimate of connectivity strength than simply analyzing the lagged correlations by means of the CCH calculated with \cref{eq:ptrans} (\cref{fig:cch_iv_triple}(b, c) dashed lines). 
We use the IV estimator given by \cref{eq:wald} on the three neuron system (\cref{fig:cch_iv_triple}(b, c) solid lines). 
It converges to the correct causal conclusions that the weights $ w_{BC} = 0.2 $ and $ w_{AC} = 0 $ as opposed to the CCH method which falsely concludes that $ w_{AC} \approx 0.1 $. 
For such a simple system, it produces meaningful estimates of the causal interactions between neurons.
\tikzset{line/.style = {draw, -latex', ultra thick}}
\begin{figure}
\makeatletter
\renewcommand\p@subfigure{}
\makeatother
\begin{subfigure}{\textwidth} \centering
\begin{tikzpicture}
    \node (u) {$ S $};
    \node [above = of u] (x) {$ A/B $};
    \node [above right = of u] (y) {$ C $};
    \node [left = of x] (sr) {$ A_r/B_r $};
    \path [line] (x) -- (y);
    \path [line] (u) -- (x);
    \path [line] (u) -- (y);
    \path [line] (sr) -- (x);
\end{tikzpicture}
\caption{} \label{fig:cch_iv_triple:0}
\end{subfigure}\medskip\\
\begin{subfigure}{\textwidth} 
\includegraphics[scale=1]{triple}
\end{subfigure}
\caption{{\bf Instrumental variable estimation (IV) of connectivity}. 
(a) during instrumental variable estimation we use a variable that is assumed to be random (here refractoriness) which influences a variable of interest (here spiking) and to use this influence to infer the causal interaction of that variable on other variables (here spiking of A or B onto C). 
A popular estimation approach for IVs, the Wald technique, correctly estimates causal connectivity in the A, B, C system using the refractory period. 
The path diagram in (a) shows the associations between neurons $ A $, $ B $ or $ C $, the stimulation $ S $, and the IV as $ A_{r} $ or $ B_{r} $ (r for refractory).
The IV estimator calculated by \cref{eq:wald} converges to $ \hat{\beta}_{BC} \approx 0.2, \hat{\beta}_{AC} \approx 0 $ after approximately $ 5000 $ trials as seen in (b, c). 
Black line indicates ground truth and the results of the cross-correlations are showed as dotted lines. 
Note the difference between the IV estimate and CC method in (c).}
\label{fig:cch_iv_triple}
\end{figure}
\FloatBarrier

>>>>

\documentclass[11pt]{article}
\usepackage{amsmath,array}
\usepackage{amsfonts}
\usepackage{amssymb}
\usepackage{amsthm}
\usepackage{breakcites}
\usepackage{graphicx,subcaption}
\usepackage{cleveref}
\usepackage{tikz}
\usepackage{placeins} %FloatBarrier
\usepackage{booktabs}
\usepackage{natbib}
\usepackage{authblk}
\usepackage{url}

\usetikzlibrary{arrows,shapes,calc}
\usetikzlibrary{positioning}

@@

 \maketitle
% suggestions for referee?
\begin{abstract} % max words 150
\noindent To study how the brain works, it is crucial to identify causal interactions between neurons, which is thought to require perturbations. However, when using optogenetics we typically perturb multiple neurons, producing a confound - any of the stimulated neurons can have affected the target. Here we show how this produces large biases, and how they can be reduced using the instrumental variable (IV) framework from econometrics. The interaction between stimulation and the absolute refractory period produces a weak, approximately random signal which can be exploited to estimate causality. When simulating integrate-and-fire neurons, we find that causal connectivity estimates from IV are better than na\"ive techniques ($R^2=0.77$ vs $R^2=0.01$). The difference is important as the estimates differ when applied to experimental data from stimulated neurons with recorded spiking activity. Our IV approach is easy to implement in the context of current experimental paradigms.

@@

\section{Introduction}
% The brain is infinitely confounded, thus correlations are not sufficient to draw conclusions about mechanisms and this is a hard problem in brain data.
The central goal of neuroscience, arguably, is to understand the mechanisms or causal chains that give rise to activity in the brain, to perception, cognition, and action. 
Complex systems such as the brain are hard to understand because of the numerous ways the contributing elements may interact internally \citep{Jonas2017}. 
Because of this, it is not sufficient to know the correlations between variables or even be able to predict them.
While observing correlations within the system is relatively easy, transitioning from observed correlations to a causal or mechanistic understanding is hard.
After all, there can be many ways that the same activities emerge from distinct causal chains \citep{drton2011global, peters2017elements}. 
Reaching a mechanistic level of understanding in the mammalian brain is incredibly hard as they contain countless neurons (e.g. 86 billion in the human brain \citep{azevedo2009equal}), each of which influences many other neurons.
Even if we could record all neurons at the same time, estimating causality and producing a mechanistic understanding would be extremely challenging.

% We use observational studies, where identifiability is questionable and causality is thus typically wrong. 
In today's typical studies, we only record from a small subset of all neurons. 
The data we obtain from such recordings, e.g. from electrophysiology or calcium imaging, is observational, which means that it does not result from randomized perturbations. 
In such cases, we can never know to which level the observed activity was caused by other observed activity, or by unobserved activity. 
The activity of the unobserved neurons is thus called confounders.
If mechanisms are estimated from observational data in the presence of confounders, the consequence may be large errors and incorrect conclusions \citep{angrist2008mostly}. 
Unobserved neural activity confounds estimates of causal interactions and makes it difficult to estimate underlying mechanisms.

% Set up confounding
Confounding is the big threat to causal validity \citep{pearl2009causality} irrespective of the use of simple regression techniques or advanced functional connectivity techniques \citep{stevenson2008inferring, honey2009predicting, aitchison2017or, pfau2013robust}.  
To estimate connectivity it is first and foremost important that the used signals reflect cause and effect, therefore we use the term causal connectivity. 
Consider an example where we want to estimate causal connectivity between two observed, but unconnected neurons, $ A $ and $ C $ (\cref{fig:intro}(a)) by stimulating optogenetically. 
A third unobserved neuron $ B $, is driven together with $ A $ by a common input and they are thus strongly correlated.
Furthermore, $ B $ and $ C $ are connected, thus the input drives $ C $ through $ B $.
Consequently, $ A $ and $ C $ are also correlated and the regression $ C = \beta A + noise$ will misleadingly conclude a direct interaction when causally interpreted. 
In this case, we say the regressor $ A $ is endogenous and the regression coefficient $ \beta $ estimates the magnitude of association rather than the magnitude of causation. 
Na\"ive regressions in partially observed systems will generally not reveal causality.

% Why do we not use GLM
A much used method for estimating the output of single neurons is to perform multiple regression analyses \citep{pillow2008spatio}, modeling each neuron with a generalized linear model (GLM). 
Multiple regression may be a solution to confounding problems as they support ``explaining away" background activity \citep{stevenson2008inferring}.
However, this is only a meaningful strategy if most neurons are included in the recordings.
Furthermore, only under certain assumptions about nonlinearity or noise sources does a fully observed system become identifiable \citep{daniusis2012inferring,shimizu2006linear}.
This is rarely the case in experimental settings, especially in the mammalian brain.
Thus, brain data, will almost never satisfy the criteria needed for identifiability \citep{pearl2009causality}. 
We thus aim to develop methods that can identify causal connectivity between neuron pairs.

% Stimulation is an intervention which is needed to estimate causal interactions
To estimate causal relationships between neurons, stimulating the pre-synaptic neuron is the gold standard. 
In fact, a common definition of causality is in terms of the effect of changing one variable in the system, independently of changing other variables -- an intervention \citep{pearl2009causality}. 
If we stimulate single neurons, the ability to estimate causal relationships by regression is within reach. 
However, this is experimentally challenging and yields low cell count because it requires intracellular, juxtacellular or two-photon stimulation \citep{pinault1996novel, lerman2017two, nikolenko2007two, emiliani2015all}. 
Because gold-standard perturbations are challenging, it is necessary and highly desirable if causality could be obtained from optogenetic stimulation in combination with neural recordings of large populations of neurons \citep{boyden2005millisecond, zemelman2002selective}.

% Optogenetic stimulation is usually totally non-local, producing confounders
Interpreting the results from optogenetic stimulation in terms of causal interactions is difficult. 
In most experimental settings, optogenetic stimulation will affect many neurons simultaneously. 
Hence, the stimulus will produce a distributed pattern of activity.
This distributed pattern of stimulation produces activity which then percolates through the network of neurons. 
Thus any post-synaptic activity induced by stimulation could in principle come from any of the stimulated neurons, introducing problematic confounders.

% Causal inference is a thing, if not done well you say stupid things
For insights into how we may resolve the confounding problem induced by optogenetic stimulation, we may look to other fields that have addressed the problem of endogenous regressors. 
The inference of causality from observational data is addressed in the fields of statistics \citep{pearl2009causality}, machine learning \citep{peters2017elements} and econometrics \citep{angrist2008mostly}. These fields have extensively worked on methods to estimate causality in the face of potential confounding and may offer us clues on how to solve our problems.

% Example
A commonly used approach towards causal inference in economics are instrumental variables. 
Let us say that we want to estimate the return $ \beta $ from education $ x $ to yearly wages $ y $ with the regression $ y = \beta x + u $. 
Here $ u $ are the factors other than education that contribute to yearly wages. 
One of the factors in $ u $ is a person's cognitive ability which may also affect education. The regressor $ x $  is thus correlated with the error term $ u $. 
This will imply that the regression estimate of $ \beta $ will not estimate the magnitude of causation from education on wages, but rather, its association. 
In this case one may use the proximity to a college or university as an instrumental variable (IV) \citep{card1993using}. 
This is following the idea that proximity to a college does not affect your cognitive ability but may affect your proclivity to attend college.
We expect living in proximity of colleges and universities to give higher probability to attend without affecting other contributing factors to wages such as cognitive ability. 
Then, in order to attribute the causal effect of education on wages one may calculate the ratio of covariances $ \beta = cov(\mathrm{proximity},\mathrm{wages})/cov(\mathrm{proximity},\mathrm{education})$. 
This ratio corrects for the confounding factor.

% Worry is this really correct? Yes, its provable. IVs are (under the right assumptions) actually correct.
The IV technique has been used extensively in econometrics and can provide provable unbiased causal estimates given three main assumptions \citep{angrist2008mostly}. 
First, the instrument must be decorrelated with the error term. 
Second, the instrument must be correlated with the regressor. 
Third, there must be no direct influence of the instrument on the outcome variable, but only an influence through the regressor variable. 
The validity of these assumptions is central when using the IV approach.

% Refractory period as an IV
For an instrument to be good, it needs to be unaffected by other variables. 
In the brain, almost everything is affected by the network state. 
However, certain variables can be more or less affected. 
For example, the overall activity of the network is due to slow and strongly nonrandom dynamics. 
In contrast, the temporal pattern of when a neuron is in a refractory state may be in good approximation random. 
First, if neurons are spiking according to conditional Poisson distributions, their exact timing conditioned on the network state, will be random. 
While refractoriness may not be perfectly random, the exact times of spiking are notoriously difficult to predict \citep{stevenson2008inferring} suggesting that refractoriness is quite random.

@@

\section{Results}
\subsection{Optogenetics is not local}
Optogenetic stimulation is generally seen as a perturbation that by-and-large affects neurons in proximity (< 1mm) of the light source with the effect decreasing with distance. 
However, this is a misleading way of conceptualizing the spatial effect of stimulation as it depends on multiple factors. 
Light intensity and opsin density are important as more light and ion channels will cause a stronger effect on each cell. 
Moreover, the number of potentially stimulated neurons is critical as more neurons will have a larger impact on the overall population activity. 
Finally, physiological properties of the cells are important as light may have a stronger effect on spiking activity when the membrane potential of the cell is sufficiently close to the firing threshold. 
The induced effect of optogenetic stimulation as a function of distance should be given by a relation between the four parameters: light intensity, spatial distribution of neurons, distributions of membrane potential across neurons, and the distribution of induced photo currents.

To estimate the light intensity, we calculated the spatial extent of laser light delivered by fiber-optics under plausible experimental conditions according to \citet{Aravanis2007}; see \cref{sec:method:opto}. 
While their experiment uses blue light, their fits assume no absorption making the equations somewhat general.
This modeling of light intensity yield an approximately $ 1/r^2 $ reduction with distance $ r $ from the stimulation site \cref{fig:concept} (cyan line). 
This is explained by the surface of a 3D shell growing with $4\pi r^2$ and photons will be roughly isotropic (but see \citep{thunemann2018does}) beyond the scattering length \cref{fig:concept} (inset). 
The same number of photons has to cross each of the spheres around the stimulation location unless they are absorbed or scattered inwards. 
As a result, the density of photons decreases with distance.

The number of illuminated neurons at a given distance will, however, increase with distance to the stimulation site given that neurons are roughly uniformly distributed in brain tissue \cref{fig:concept} (black line). 
In fact, it will increase by approximately $ r^2 $ with distance. 
This derives from the same surface scaling as for the 3D shell as for the photon flow. 
Thus the number of neurons that can be activated increases rapidly with distance.

To estimate the effect of stimulation, the mechanism with which light affects spiking activity needs to be considered. 
This can largely be characterized by the  distribution of membrane potentials across neurons. 
Surprisingly, this distribution has been observed to be symmetrically distributed and relatively flat \citep{pare1998impact,destexhe1999impact,rudolph2006use}. 
The expected response from a pulse of light that induces a charge $Q$ should be proportional to the number of affected neurons whose membrane potential sit within a $Q/C$ range of the threshold ($C$ is the capacitance). 
Given that the distribution of membrane potentials is relatively flat (the density close to the threshold is generally within an order of magnitude of the density of its mode) suggests that the spiking response to a perturbation for any neuron is roughly proportional to the induced photo current. 
The peak amplitude of the photo current relates approximately logarithmically to the light intensity \citep{wang2007high}; see \cref{sec:method:opto}. 
Assuming that opsins are evenly distributed across neurons, the induced photo current will not be proportional to light intensity - it will fall slower. 
Based on this, we calculate the overall stimulation effect to be the product of the number of neurons in a spherical slice and the peak amplitude photo current. 
This product actually increases with distance (up to the distance where absorption becomes important) \cref{fig:concept} (blue line). In other words, there is more activation at 500um than at 100um.
Thus, optogenetic stimulation utilizing single photon activation does not produce a localized effect.

\begin{figure}\includegraphics[scale=.95]{opto-powerlaw}
\caption{{\bf Spatial extent of optogenetic stimulus}. 
Due to scattering and geometric loss the light intensity (I, cyan line) with an intensity of $10 mW/mm^2$ exiting the optogenetic fiber follows approximately an inverse square law $ r^{-2} $ where $ r $ is the distance from the fiber. 
If neurons are uniformly distributed, the number of affected neurons in a spherical slice increases by $ r^{2} $ (N, black line). 
The total photo current (P, blue line) calculated as the sum over neurons of peak amplitude photo current in a spherical slice thus increases with distance due to the nonlinear relation between light intensity and photo current, depicted as percentage of maximum.}
\label{fig:concept}
\end{figure}

@@

After running the simulation, the peri-stimulus time histogram of the stimulated neurons (\cref{fig:intro}(b)) shows the result of both the stimulation itself (suppressed for visibility) and the neuron's refractory period \cref{fig:intro}(c) (AA, BB).
Since the stimulation affects A and B simultaneously, it induces a strong correlation between A and B \cref{fig:intro}(c) (AB). 
This further generates a strong correlation between A and C, confounding the system by rendering the cross-correlation histograms (CCHs) between BC and AC both statistically significant ($ p_{\mathrm{fast}} < 0.001, p_{\mathrm{diff}} < 0.001 $; see \cref{sec:method:cch}). 
A na\"ive reading of this result may suggest causal influences of both A and B on C.

Even though the correlation peak between B and C is larger than between A and C, due to correlated spikes outside the periods with stimulation, one may imagine a situation where only A and C is measured, giving rise to a false prediction that they are connected. 
Alternatively, A may have a stronger response to the optogenetic stimulation, in which case we may even have a stronger A-C correlation than B-C correlation.  
If stimulation affects multiple neurons simultaneously, there is a real confounding problem.

\begin{figure} 
\makeatletter
\renewcommand\p@subfigure{}
\makeatother
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{simple}
\caption{} \label{fig:intro:1}
\end{subfigure}\hfill
\begin{subfigure}{0.485\textwidth} \includegraphics[scale=1]{psth_triple}
\caption{} \label{fig:intro:2}
\end{subfigure}\\
\begin{subfigure}{\textwidth}\centering \includegraphics[scale=1]{xcorr}
\caption{} \label{fig:intro:3}
\end{subfigure}
\caption{{\bf Optogenetic stimulation induces spurious correlations}. 
Simple network containing three neurons shows stimulation configuration with blue laser light and the connections with arrows (a). 
The neurons A and B are stimulated in 1000 trials and the corresponding peristimulus time-histogram are shown in (b) upper panel with a raster plot in the lower panel. 
Cross-correlation histograms (CCHs) are shown in (c) where axes represent time lag in ms and counts of coincident spikes in bins of 1 ms.}
\label{fig:intro}
\end{figure}

\FloatBarrier
\subsection{Instrumental variables to resolve confounding}
In order to estimate the actual influence of stimulation of a neuron on post-synaptic neurons, we need to distinguish the influence of one stimulated neuron from the influence of another stimulated neuron. 
We would thus need something that affects the stimulation effect separately across neurons. 
Arguably, refractoriness is such a variable. 
If a neuron is in its absolute refractory period, then no amount of stimulation will make it spike. 
This gives us an interesting way of inferring causality, by comparing the network state between a time when a neuron is able to spike and when the neuron is unable to spike when the stimulation hits.

Instrumental variables require the existence of a variable (e.g. refractoriness) that affects one variable of interest (the presynaptic neuron), but affects the rest of the network (including the postsynaptic neuron) only through that variable. 
This independent influence then allows quantifying the influence of the variable of interest on the rest of the network. 
In our case, the refractory states of a neuron is in good approximation independent on short time scales (see Discussion for caveats). 
It affects the influence of stimulation on the pre-synaptic neuron (\cref{fig:cch_iv_triple}(a)). 
The trials where a stimulus is unable to elicit a spike due to the refractory state can then be used to identify causal effect on the putative post-synaptic neuron.

We can now investigate if the use of an instrumental variable gives a better estimate of connectivity strength than simply analyzing the lagged correlations by means of the CCH calculated with \cref{eq:ptrans} (\cref{fig:cch_iv_triple}(b, c) dashed lines). 
We use the IV estimator given by \cref{eq:wald} on the three neuron system (\cref{fig:cch_iv_triple}(b, c) solid lines). 
It converges to the correct causal conclusions that the weights $ w_{BC} = 0.2 $ and $ w_{AC} = 0 $ as opposed to the CCH method which falsely concludes that $ w_{AC} \approx 0.1 $. 
For such a simple system, it produces meaningful estimates of the causal interactions between neurons.

\begin{figure}
\tikzset{line/.style = {draw, -latex', ultra thick}}
\makeatletter
\renewcommand\p@subfigure{}
\makeatother
\begin{subfigure}{\textwidth} \centering
\begin{tikzpicture}
    \node (u) {$ S $};
    \node [above = of u] (x) {$ A/B $};
    \node [above right = of u] (y) {$ C $};
    \node [left = of x] (sr) {$ A_r/B_r $};
    \path [line] (x) -- (y);
    \path [line] (u) -- (x);
    \path [line] (u) -- (y);
    \path [line] (sr) -- (x);
\end{tikzpicture}
\caption{} \label{fig:cch_iv_triple:0}
\end{subfigure}\medskip\\
\begin{subfigure}{\textwidth} 
\includegraphics[scale=1]{triple}
\end{subfigure}
\caption{{\bf Instrumental variable estimation (IV) of connectivity}. 
(a) during instrumental variable estimation we use a variable that is assumed to be random (here refractoriness) which influences a variable of interest (here spiking) and to use this influence to infer the causal interaction of that variable on other variables (here spiking of A or B onto C). 
A popular estimation approach for IVs, the Wald technique, correctly estimates causal connectivity in the A, B, C system using the refractory period. 
The path diagram in (a) shows the associations between neurons $ A $, $ B $ or $ C $, the stimulation $ S $, and the IV as $ A_{r} $ or $ B_{r} $ (r for refractory).
The IV estimator calculated by \cref{eq:wald} converges to $ \hat{\beta}_{BC} \approx 0.2, \hat{\beta}_{AC} \approx 0 $ after approximately $ 5000 $ trials as seen in (b, c). 
Black line indicates ground truth and the results of the cross-correlations are showed as dotted lines. 
Note the difference between the IV estimate and CC method in (c).}
\label{fig:cch_iv_triple}
\end{figure}
\FloatBarrier
